`treats`: a modular `R` package for simulating trees and traits.

### 2023-12-15 paper/treats_paper.Rmd
Updated paper with response
Needs final check before resubmit

---
title: "`treats`: a modular `R` package for simulating trees and traits."
author: "Thomas Guillerme$^1$"
bibliography: References.bib
# csl: mee.csl
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    keep_tex: yes
    fig_width: 8
    fig_height: 8
  bookdown::html_document2: default
  bookdown::pdf_document2: default
---

@@

```{r prepdata, echo = TRUE, fig.cap="Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset. The red line represents the K-Pg boundary. Is this change in disparity related to the K-Pg mass extinction? Or, more generally, is it even possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
## Plotting the tree and the disparity through time
par(mfrow = c(1,2), bty = "null")
## Thre tree
plot(BeckLee_tree, show.tip.label = FALSE)
axisPhylo()
abline(v = BeckLee_tree$root.time - 66, col = "red")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)

@@

The first and simplest way is to simulate tree topologies that have similar properties than the observed one.
To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
These two parameters can be crudely calculated from a given tree by counting the number of speciations and extinctions through time (but see better methods, e.g. `ape::birthdeath`; @lawton1995extinction; @ape).
We also need a stopping rule for when to stop the simulations (in our case when reaching 30 time units).
This will produce a single random tree using the input parameters (Fig. 3).
The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
Using 30 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data.
Note that I will not discuss the options here in great details.
Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).

```{r, message = FALSE}
## Using the birth-death parameters from the observed tree
my_bd_params <- crude.bd.est(BeckLee_tree)
## Setting the stopping rule (stop after 30 time units)
stop_rule <- list(max.time = 30)

## Simulate the tree
set.seed(1)
sim_tree <- treats(bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100,
                   verbose    = FALSE)
```
```{r simtree, echo = FALSE, fig.cap="A randomly simulated tree with similar properties as the observed one. Note that the time here is expressed in arbitrary units.", fig.width = 6, fig.height = 6}
## Plotting the resulting tree
plot(sim_tree, show.tip.label = FALSE)
## Adding the phylogenetic axis
axisPhylo()
```

@@

## Simulate the tree and traits
set.seed(1)
sim_data <- treats(traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100,
                   verbose    = FALSE)
```
```{r simtreetraits, echo = TRUE, fig.cap="A randomly simulated tree with a 2 dimensional random Brownian Motion trait. Orange, light blue and dark blue dots respectively represent nodes, fossils and living species.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Plotting one trait through time
plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")

@@

To answer our question, we also want to simulate an extinction event.
To do so, we can create two different `"events"` object with the function `make.events`.
The first one will simulate a random extinction event after reaching half of the simulations (15 time units) and then making three quarter (0.75) of the taxa go extinct.
The second one will simulate a random extinction but based on trait values: after reaching time 15, all the species with positive trait values will go extinct.
Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e. half the species are expected to evolve a trait value above 0).
For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex.
Because of this stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic (Fig. 5).

```{r, message = FALSE}
## Creating a random mass extinction
random_extinction <- make.events(
    target       = "taxa",
    condition    = time.condition(15),
    modification = random.extinction(0.75))
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target = "taxa",
    condition = time.condition(15),
    modification = trait.extinction(x = 0, condition = `>=`))

set.seed(1)
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a selective extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = positive_extinction,
                   null.error = 100,
                   replicates = 50)
```
```{r simtreetraitsevents, echo = FALSE, fig.cap="Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct. The red line marks the extinction event.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Visualising the difference between both scenarios
## Random extinction
plot(drop.singles(sim_rand_extinction[[31]]), main = "Random extinction")
abline(v = 15, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)

## Selective extinction
plot(drop.singles(sim_trait_extinction[[41]]), main = "Positive extinction")
abline(v = 15, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
```

@@

## Calculate the dispRity for all the simulations
random_extinction_disparity <- dispRitreats(sim_rand_extinction,
                                            method = "continuous",
                                            model  = "proximity",
                                            time   = seq(from = 25, to = 0, by = -5),
                                            metric = c(sum, variances),
                                            scale.trees = FALSE)
selective_extinction_disparity <- dispRitreats(sim_trait_extinction,
                                            method = "continuous",
                                            model  = "proximity",
                                            time   = seq(from = 25, to = 0, by = -5),
                                            metric = c(sum, variances),
                                            scale.trees = FALSE)

## Scale the disparity results (to compare to the observed ones)
random_extinction_disparity <- rescale.dispRity(random_extinction_disparity)
selective_extinction_disparity <- rescale.dispRity(selective_extinction_disparity)
observed_disparity <- unlist(get.disparity(rescale.dispRity(observed_disparity)))

## Plotting the results with the observed disparity
par(mfrow = c(1,2))
plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances")
abline(v = 3, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
plot(selective_extinction_disparity, main = "Selective extinction", ylab = "")
abline(v = 3, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
```

@@

## Package location
The `treats` package is available 
<!-- % on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or  -->
on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).

>>>>

---
title: "`treats`: a modular `R` package for simulating trees and traits."
author: "Thomas Guillerme$^1$"
bibliography: References.bib
# csl: mee.csl
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    keep_tex: yes
    fig_width: 8
    fig_height: 8
  # bookdown::html_document2: default
  bookdown::pdf_document2: default
---

@@

```{r prepdata, echo = TRUE, fig.cap="Observed phylogeny (A) and disparity through time (B) in the subset of @beck2014ancient's dataset. The red line represents the K-Pg boundary. Is this change in disparity related to the K-Pg mass extinction? Or, more generally, is it even possible to detect potential changes of disparity due to a mass extinction event?", fig.width = 18, fig.height = 8}
## Plotting the tree and the disparity through time
par(mfrow = c(1,2), bty = "null")
## Thre tree
plot(ladderize(BeckLee_tree), show.tip.label = FALSE)
axisPhylo()
abline(v = BeckLee_tree$root.time - 66, col = "red")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)

@@

The first and simplest way is to simulate tree topologies that have similar properties than the observed one.
To do so, we need to use some **speciation** parameter indicating the rate at which lineages speciate (_aka_ "birth" or "$\lambda$") and an **extinction** parameter indicating the rate at which they go extinct (_aka_ "death" or "$\mu$").
Here we are using two relatively arbitrary (speciation = 0.035 and extinction = 0.02) to get trees roughly matching the observed tree.
Note that you might want to consider more appropriate ways to calculate these rates for research projects (e.g @magallon2001absolute).
We also need a stopping rule for when to stop the simulations (in our case when reaching 140 time units).
This will produce a single random tree using the input parameters (Fig. 3).
The number of time units in `treats` is arbitrary and is not equivalent to millions of years.
Using 140 time units here allows to simulate number of tips in a similar order of magnitude as the ones in the observed data.
Note that I will not discuss the options here in great details.
Much more information can be found in the [`treats` manual](http://tguillerme.github.io/treats.html).

```{r, message = FALSE}
## Using the birth-death parameters from the observed tree
my_bd_params <- make.bd.params(speciation = 0.035, extinction = 0.02)
## Setting the stopping rule (stop after 140 time units)
stop_rule <- list(max.time = 140)

## Simulate the tree
set.seed(2)
sim_tree <- treats(bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100,
                   verbose    = FALSE)
```
```{r simtree, echo = FALSE, fig.cap="A randomly simulated tree with similar properties as the observed one. Note that the time here is expressed in arbitrary units.", fig.width = 6, fig.height = 6}
## Plotting the resulting tree
plot(ladderize(sim_tree), show.tip.label = FALSE)
## Adding the phylogenetic axis
axisPhylo()
```

@@

## Simulate the tree and traits
set.seed(123)
sim_data <- treats(traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   null.error = 100,
                   verbose    = FALSE)
```
```{r simtreetraits, echo = TRUE, fig.cap="A randomly simulated tree with a 2 dimensional random Brownian Motion trait. Orange, light blue and dark blue dots respectively represent nodes, fossils and living species.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Plotting one trait through time
plot(sim_data, ylab = "Trait 1", las = 1, main = "Trait 1 through time")

@@

To answer our question, we also want to simulate an extinction event.
To do so, we can create two different `"events"` object with the function `make.events`.
The first one will simulate a random extinction after reaching 66 time units and then making three quarter (0.75) of the taxa go extinct.
The second one will simulate a random extinction but based on trait values: after reaching time 66, all the species with positive trait values will go extinct.
Both scenarios illustrate two different types of mass extinctions but they are not equivalent: because of the ancestral trait value starting at 0, we expect the second scenario to remove on average only 50% of the species (i.e. half the species are expected to evolve a trait value above 0).
For more details on simulating the effect of mass extinction and the difficulties to simulate an unambiguous effect of a mass extinction, see @puttick2020complex.
Because of this stochasticity of the simulations, we will repeat them 50 times (using `replicates = 50`) to generate a distribution of possible simulated scenarios as opposed to a random single one that could be idiosyncratic (Fig. 5).

```{r, message = FALSE}
## Creating a random mass extinction
random_extinction <- make.events(
    target       = "taxa",
    condition    = age.condition(140-66),
    modification = random.extinction(0.75))
## Creating an extinction that removes species with positive trait values
positive_extinction <- make.events(
    target = "taxa",
    condition = age.condition(140-66),
    modification = trait.extinction(x = 0, condition = `>=`))

set.seed(123)
## Simulate the tree and traits with a random extinction event
sim_rand_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = random_extinction,
                   null.error = 100,
                   replicates = 50)

## Simulate the tree and traits with a selective extinction event
sim_trait_extinction <- treats(
                   traits     = my_traits,
                   bd.params  = my_bd_params,
                   stop.rule  = stop_rule,
                   events     = positive_extinction,
                   null.error = 100,
                   replicates = 50)
```
```{r simtreetraitsevents, echo = FALSE, fig.cap="Two different mass extinction events: (A) 75% of species go extinct; (B) all species with positive trait values go extinct. The red line marks the extinction event.", fig.width = 16, fig.height = 8}
par(mfrow = c(1,2))
## Visualising the difference between both scenarios
## Random extinction
plot(sim_rand_extinction[[42]], main = "Random extinction")
abline(v = 66, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "A", bty = "n", cex = 2)
## Selective extinction
plot(sim_trait_extinction[[50]], main = "Positive extinction")
abline(v = 66, col = "red")
abline(h = 0, col = "grey")
legend("topleft", pch = NULL, legend = "B", bty = "n", cex = 2)
```

@@

## Calculate the dispRity for all the simulations
random_extinction_disparity <- dispRitreats(sim_rand_extinction,
                                            method = "continuous",
                                            model  = "proximity",
                                            time   = seq(from = 96, to = 36, by = -10),
                                            metric = c(sum, variances),
                                            scale.trees = FALSE)
selective_extinction_disparity <- dispRitreats(sim_trait_extinction,
                                            method = "continuous",
                                            model  = "proximity",
                                            time   = seq(from = 96, to = 36, by = -10),
                                            metric = c(sum, variances),
                                            scale.trees = FALSE)

## Scale the disparity results (to compare to the observed ones)
random_extinction_disparity <- scale.dispRity(random_extinction_disparity)
selective_extinction_disparity <- scale.dispRity(selective_extinction_disparity)
observed_disparity <- unlist(get.disparity(scale.dispRity(observed_disparity)))

## Plotting the results with the observed disparity
par(mfrow = c(1,2))
plot(random_extinction_disparity, main = "Random extinction", ylab = "Scaled sum of variances", ylim = c(0, 1))
abline(v = 4, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
plot(selective_extinction_disparity, main = "Selective extinction", ylab = "", ylim = c(0, 1))
abline(v = 4, col = "red")
lines(x = 1:7, y = observed_disparity, lty = 2, lwd = 2)
```

@@

## Package location
The `treats` package is available on the [CRAN](https://cran.r-project.org/web/packages/treats/index.html) or on [GitHub](https://github.com/TGuillerme/treats) with more associated information.
All the versions of the package are archived on ZENODO with associated [DOI:10.5281/zenodo.7970384](https://doi.org/10.5281/zenodo.7970384).
